---
title: "Principal component analysis - the tidyverse way"
author: "Pietro Delfino"
date: '2018-09-05'
tags:
- R coding
- statistics
- PCA
- tidyverse
categories: R, github, stats, tidyverse, PCA
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(skimr)
```

# PCA with any kind of data

In my daily work I am often asked to perform [**Principal Component Analysis**](https://en.wikipedia.org/wiki/Principal_component_analysis) of various type of *-omics* data, for example data coming from RNA-Seq.

With this data usually the rows correspond to genes (transcripts, proteins, etc..) while columns are individuals or samples.

For example something like this, where I have mouse proteins in rows and mass spec peak intensities in columns, where each column is a different sample:

```{r, include=FALSE, error=FALSE, warning=FALSE}
x <- read_xlsx('QQ colture dati aggiuntivi 28082018.xlsx')

x <- 
    x %>%
    filter(!is.na(Row))

names(x) <- gsub('Coltura organotipica ', '', names(x))
names(x) <- gsub('TOPO_', '', names(x))
names(x) <- gsub('SWATH', '', names(x))
```

```{r echo=FALSE, error=FALSE, warning=FALSE}
x %>% 
    select(`Peak Name`, 8:11)
```

I recently started the transition to the [**tidyverse**](https://www.tidyverse.org/) approach of data science into R, after having learnt R the classic way a couple of years ago.

Also for performig PCA I came up with a tidyverse script that goes from a dataframe (a tibble) like the one above and the output of a PCA in a couple of steps.

Before going to the PCA step I want to show how I usually proceed with exploratory analysis on a tibble like this.

First thing I usually do is a `glimpse` of the tibble that gives me an overview of all the variables in my df.

```{r}
glimpse(x)
```

We can see that the values that I'd need for the PCA are in columns of the double type, but we also have two columns, `Row` and `Index` which are of the same double type. Keep this in mind for later. For computing PCA I only need the columns with the mass spec intensities values, then I want to focus on those columns, using another wonderful command from the `skimr` package.

```{r}
skim(x[8:length(names(x))])
```

With the `skimr` command I see that all the variables are complete, so there are no missing values, which is good for PCA, that usually does not like `NA`'s.

Why not doing also a boxplot?

```{r, dev='svg'}
x %>% 
    select(matches('ug')) %>%
    rownames_to_column %>%
    gather(variable, value, -rowname) %>%
    ggplot(aes(variable, log2(value + 1))) + geom_boxplot() +
    coord_flip() +
    theme_bw()
```



