---
title: "Customize ggplot2 boxplot - make it beautiful"
author: "Pietro Delfino"
date: '2019-07-15'
tags:
- R coding
- ggplot
- statistics
categories:
- R
- stats
- ggplot
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(reshape2)
```

I like `ggplot2` graphic style, but to make publication acceptable plots, especially to align with graphic style of some journals, sometimes it is necessary to customize plots and modify some parameters that usually are kept as default.

Here I will present some tricks to make beautiful - at least to me - boxplots. I am using the iris dataset, always the best to be reproducible for everyone.

First we define a function to make first letter of each word upper case, retrieved from here https://stackoverflow.com/a/18509816/5575616.


```{r}
firstup <- function(x) {
    substr(x, 1, 1) <- toupper(substr(x, 1, 1))
    x
}
```

I also make use of a nice package that I use almost every time I want different and original colors for my plots. `ggsci` (https://cran.r-project.org/web/packages/ggsci/vignettes/ggsci.html).

Now the plot, then we comment.

```{r message=FALSE, warning=FALSE, dev='svg',out.width='120%'}
library(tidyverse)
library(reshape2)
library(ggsci)

iris %>% 
    as_tibble() %>% 
    melt() %>% 
    mutate(Species=as.character(Species) ) %>% 
    mutate(Species=firstup(Species) ) %>% 
    dplyr::filter(variable == 'Sepal.Length') %>% 
    
    ggplot(aes(Species, value, fill=Species)) +
    geom_boxplot(outlier.shape = NA) +
    
    coord_flip() +
    annotate('rect', xmin = 0.5, xmax = 1.5, ymin = -Inf, ymax = Inf,fill = "grey", alpha = 0.2) +
    annotate('rect', xmin = 2.5, xmax = 3.5, ymin = -Inf, ymax = Inf,fill = "grey", alpha = 0.2) +
    
    scale_y_continuous(expand = c(0, 0), 
                       limits = c(floor(min(iris$Sepal.Length)),
                                  ceiling(max(iris$Sepal.Length)))) +
    
    geom_boxplot(outlier.shape = NA) +
    
    theme(axis.line.y = element_blank(),
          axis.line.x = element_line(),
          panel.grid = element_blank(),
          axis.text.y = element_text(size = 16, color='black'),
          axis.text.x = element_text(size = 10, color='black'),
          axis.ticks.y = element_blank(),
          axis.title.y = element_blank(), 
          axis.title.x = element_text(size = 16, color='black'),
          panel.background = element_blank(),
          axis.ticks.length = unit(.25, "cm"),
          legend.position = 'none') +
    ylab('Sepal Length') +
    ggsci::scale_fill_npg()
```

The main point here is to create an alternate background with custom colors that makes the reading of the plot more pleasant.

The `ggplot2` function `annotate` is a useful proxy for this behaviour. The trick here is to call again the main plot object (in this case `geom_boxplot`) after the `annotate` function, otherwise it remains covered.

Another interesting custom trick is to use the `scale_y_continuous` function in combination with `expand` to make the ticks start and end on a custom value, in this case from 4 and 8; a behaviour that is not default in `ggplot2`.

All the options within the `theme` function are modified to remove completely the panel grid and background.

Now I am working to find a way to automatically set the annotate boxes limits in case of multiples (more than 3), and not known in advance, variables.

