---
title: "Customize ggplot2 boxplot - make it beautiful"
author: "Pietro Delfino"
date: '2019-07-15'
tags:
- R coding
- ggplot
- statistics
categories:
- R
- stats
- ggplot
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(reshape2)
```

I like `ggplot2` graphic style, but to make publication acceptable plots, especially to align with graphic style of some journals, sometimes it is necessary to customize plots and modify some parameters that usually are kept as default.

Here I will present some tricks to make beautiful - at least to me - boxplots. I am using the iris dataset, always the best to be reproducible for everyone.

First we define a function to make firs letter of each word upper case, retrieved from here https://stackoverflow.com/a/18509816/5575616.


```{r}
firstup <- function(x) {
    substr(x, 1, 1) <- toupper(substr(x, 1, 1))
    x
}
```

We also make use of a nice package that I use almost every time I want different and original colors for my plots. `ggsci` (https://cran.r-project.org/web/packages/ggsci/vignettes/ggsci.html).

Now the plot, then we comment.

```{r message=FALSE, warning=FALSE, dev='svg'}
library(tidyverse)
library(reshape2)
library(ggsci)

iris %>% 
    as_tibble() %>% 
    melt() %>% 
    mutate(Species=as.character(Species) ) %>% 
    mutate(Species=firstup(Species) ) %>% 
    
    ggplot(aes(Species, value, fill=Species)) +
    geom_boxplot(outlier.shape = NA) +
    
    coord_flip() +
    annotate('rect', xmin = 0.5, xmax = 1.5, ymin = -Inf, ymax = Inf,fill = "grey", alpha = 0.2) +
    annotate('rect', xmin = 2.5, xmax = 3.5, ymin = -Inf, ymax = Inf,fill = "grey", alpha = 0.2) +
    
    scale_y_continuous(expand = c(0, 0), limits = c(0,8)) +
    
    geom_boxplot(outlier.shape = NA) +
    
    theme(axis.line.y = element_blank(),
          axis.line.x = element_line(),
          panel.grid = element_blank(),
          axis.text.y = element_text(size = 10, color='black'),
          axis.text.x = element_text(size = 10, color='black'),
          axis.ticks.y = element_blank(),
          axis.title.y = element_blank(), 
          panel.background = element_blank(),
          axis.ticks.length = unit(.25, "cm"),
          legend.position = 'none') +
    ylab('Length') +
    ggsci::scale_fill_npg()
```

The main point here is to create an alternate background with custome color that makes the reading of the plot more pleasant.

The `ggplot2` function `annotate` is a useful proxy for this behaviour. The trick here is to call again the main plot object (in this case `geom_boxplot`) after the `annotate` function, otherwise it remains covered.

Another interesting custom trick is to use the `scale_y_continuous` function in combination with `expand` to make the axis start from a custom value, in this case from zero; a behaviour that is not default in `ggplot2`.



