---
title: "Gene expression - microarray data still useful"
author: "Pietro Delfino"
date: '2019-07-17'
tags:
- R coding
- ggplot
- statistics
- microarray
- gene expression
- bioinformatics
categories:
- R
- stats
- ggplot
- bioinformatics
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(reshape2)
library(GEOquery)
library(DT)
library(knitr)
```

I work on pancreatic cancer **-omics**. Data mining on published data is a real source of gold today when someone wants to find confirmation of an hypothesis on a more general scale than his own dataset.

The most common -omics datasets out there are transcriptomics one; both RNA-Seq and good ol' microarray. Regarding RNA-Seq, the well known TCGA and ICGC are my starting point for any explorative data mining analysis.

Regarding microarray, few studies with clear data have been published in the past, and someone has already collected the most interesting ones (see [Birnbaum et al. 2017](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5674743/), [Zhao et al. 2018](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5975421/)).

For the scopes of my research I have also explored some of these datasets. The majority of theme are stored in the [GEO database](https://www.ncbi.nlm.nih.gov/geo/).

Here I want to present a basic workflow to retrieve a microarray expression dataset from GEO and to perform basic explorative analysis in ***R***. As an example I take [GSE62165](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE62165) dataset.

To download and access data I use the *Bioconductor* `GEOquery` package (https://bioconductor.org/packages/release/bioc/html/GEOquery.html).

```{r message=FALSE, warning=FALSE, dev='svg',eval=FALSE}
# Load some libraries
library(tidyverse)
library(reshape2)
library(ggsci)
library(GEOquery)

# Downlaod ExpressionSet
x <- getGEO('GSE62165', destdir = 'your/favourite/path', getGPL = T)
```

We have downloaded the *GSE series matrix* to file and the *GPL .soft annotation*.
Now we load them back.

```{r message=FALSE, warning=FALSE, dev='svg', eval=FALSE}
x <- getGEO(filename = 'your/favourite/path/GSE62165_series_matrix.txt.gz')
```

```{r message=FALSE, warning=FALSE, dev='svg', eval=FALSE}
x <- getGEO(filename = 'GSE62165_series_matrix.txt.gz')
```

Then we can start exploring the data

```{r message=FALSE, warning=FALSE, dev='svg', results='asis'}
# First we save the expression matrix in a different object
#mat <- exprs(x)

#dim(mat)

#kable(mat[1:4, 1:8], caption = 'Firt line of the expression matrix')

#boxplot(mat, outline=F)
```

